---
title: "R tutorial medicine"
author: "Aayush Kapri"
date: "2023-09-12"
output:
  pdf_document: default
  html_document: default
warning: FALSE
---
## 1 **R and RStudio**

### 1.1 **R**

R is a programming language which is highly relevant in medicine and health science due to its robust data analysis and statistical capabilities. Researchers in these fields  can use R to analyze large datasets, conduct statistical modeling, and visualize medical data. It is open-source and has extensive libraries, which make it a powerful tool for advancing medical research and improving healthcare practices.

To install R, go to [cran](https://cran.r-project.org/). And download the version that is appropriate to your device. It is recommended to install the latest version of R. If you already have R installed in your device, you can update your current version by installing the latest version. You do not need to unistall the previous version to update your previous version. 

### 1.2 **RStudio**

RStudio is an integrated development environment (IDE) for the R programming language. RStudio provides a user-friendly and feature-rich interface for working with R, making it easier  to write, debug, and run R code efficiently.

In order to install free version of RStudio, go to [the download page](https://posit.co/download/rstudio-desktop/#download). You need to download the version that is appropriate to your device. 

### 1.3 **RStudio Cloud**

R programming can also to used in a cloud version of RStudio, making users to access R and their work from anywhere with a web browser. Visit [Postit](https://posit.cloud/) and it will have the exact interface as Rstudio where you can write codes and it will be stored in the cloud. This platform facilitates collaboration between researchers by enabling simultaneous workflows.

## 2 **Operators, Variables and Data Types **

### 2.1 **Operators**

Before proceeding further, let us introduce the operator that is most commonly used in R, i.e., '<-'. It is the assignment operator.

```{r}
# Suppose we have a patient's age and want to store it in a variable.
patient_age <- 35

# Now, you can use 'patient_age' for medical calculations or analysis.
# print() is used for printing in R. \n is used to break lines.
print(paste("The patient's age is:", patient_age, "years old"))

```
Operators in R should be very familiar to researchers. 

``` {r}
# Example 1: Addition (+)
dose_1 <- 100
dose_2 <- 50
total_dose <- dose_1 + dose_2
print(paste("Total dose:", total_dose, "mg"))

# Example 2: Subtraction (-)
pre_treatment_bp <- 120
post_treatment_bp <- 110
delta_bp <- pre_treatment_bp - post_treatment_bp
print(paste("Change in blood pressure:", delta_bp, "mmHg"))

# Example 3: Multiplication (*)
weight_kg <- 70
height_m <- 1.75
bmi <- weight_kg / (height_m * height_m)
print(paste("BMI:", bmi))

# Example 4: Division (/)
total_beats <- 4500
time_in_minutes <- 60
heart_rate <- total_beats / time_in_minutes
print(paste("Heart rate:", heart_rate, "beats per minute"))

# Example 5: Exponentiation (^ or **)
base_dosage <- 10
weight_kg <- 65
drug_dosage <- base_dosage * (weight_kg^2)
print(paste("Drug dosage:", drug_dosage, "mg"))


```
### 2.2 **Variables Types**

Variables in R are named storage containers that hold data or values, allowing you to work with and manipulate information in your R scripts and programs. These variables can represent a wide range of data types, including numeric values, text, logical values, factors, and dates. You can create, assign, and modify variables to perform various data analysis and calculations in R.


```{r}
# Numeric (for representing measurements):
# Example: Patient's Body Mass Index (BMI)
bmi_numeric <- 25.5
bmi_numeric

# Integer (for whole numbers):
# Example: Patient's Age
age_integer <- 35L
print(paste("Age (integer):", age_integer))


# Character (for textual data or patient names):
# Example: Patient's Name
patient_name_char <- "John Doe"
patient_name_char

# Logical (for binary data, like whether a patient is a smoker or not):
# Example: Smoking Status
smoker_logical <- TRUE  # If the patient is a smoker
smoker_logical

# Factor (for categorical data like blood types):
# Example: Blood Type
blood_type_factor <- factor("A", levels = c("A", "B", "AB", "O"))
blood_type_factor


# Date (for recording dates of medical events):
# Example: Date of Admission
admission_date_date <- as.Date("2023-09-14")
admission_date_date


```
### 2.3 **Data Format**
In R, there exists a diverse range of data types, including scalars, vectors (both numeric, character, and logical), matrices, data frames, lists, etc. We consider the most important data type to be the data frame, for which we provide a separate section. Examples of other data types are given below. 
```{r}
# Vectors (for storing a sequence of data):
# Example: Heart rate measurements over time.
heart_rate_vector <- c(72, 75, 80, 76, 82)
heart_rate_vector

# Matrices (two-dimensional arrays with rows and columns):
# Example: Blood pressure measurements of patients over time.
patient_bp_matrix <- matrix(c(120, 130, 115, 122, 135, 118), nrow = 3, ncol = 2)
patient_bp_matrix


# Lists (for storing heterogeneous data types):
# Example: A list of patient characteristics and test results.
patient_info_list <- list(
  Name = "John Doe",
  Age = 35,
  Diagnosis = "Hypertension",
  TestResults = c(120, 75, 160)
)
patient_info_list


```
#### 2.3.1 **Data Frame**

Data frames in R are similar to tables of data, akin to spreadsheets in Excel. Unlike matrices, data frames can accommodate columns containing categorical, logical, numerical, and other types of data. When we need to import an external file, often in a tabular format, we typically load it as a data frame in R.  

```{r}
# Create a data frame for patient information
patient_data <- data.frame(
  PatientID = c(1, 2, 3, 4, 5),
  Name = c("John Doe", "Alice Smith", "Bob Johnson", "Emily Davis", "Michael Wilson"),
  Age = c(45, 28, 35, 52, 40),
  Gender = c("Male", "Female", "Male", "Female", "Male"),
  Diagnosis = c("Hypertension", "Diabetes", "Asthma", "Heart Disease", "Allergies"),
  BloodPressure = c("120/80", "135/88", "122/78", "140/95", "118/72")
)

# Display the patient data frame
print(patient_data)

```

## 4 **Packages and Libraries**

Packages in R are used to extend the functionality of the programming ability of R. Packages allow programmers to perform specialized task such as data analysis, visualization, statistical modeling and more. Popular packages in R include ```dplyr```and ``` ggplot2 ```. In order to install packages, run ```install.packages("ggplot2")``` at the command line. That will have R automatically download the package to your computer. Similarly, you can install other packages. 

### 4.1 **tidyverse, survival and ggplot**

The ```tidyverse``` is a library with collection of R packages designed for data science that share a common philosophy and grammar, making data manipulation and visualization in R more consistent and efficient. It includes popular packages like ```ggplot2```, ```dplyr```, ```tidyr```, ```readr```, and more. 

``` {r}
# Assuming lung_data is your dataset
library(tidyverse)

library(survival)
#data function can display all available data sets in this package
data(package = "survival")
# We will use data set "lung

#the sex variable in a dataset is coded as 1 for male and 2 for female, and we want to convert it to "Male" and "Female," we can use the mutate function from dplyr
lung %>%
  mutate(sex = ifelse(sex == 1, "Male", "Female")) %>% 
  head()



```
### 4.2 **Pipes**

You may have noticed or notice the operator ```%>%```. It is called the pipe operator. It is from the ```magittr``` package. Pipes are a powerful tool for enhancing the readability and expressiveness of your code. It allows you to write code in a way that flows from left to right, making it easier to understand. As ```%>%``` is a bit tedious to type, there exist shortcuts: shift-ctrl-m on a Mac, shift-ctrl-m on a Windows machine. Here is an example of code with and without pipe:

```{r}
# Load the necessary library
library(survival)

# Load datset
data(cancer, package = "survival")

# Without pipes
male_data <- subset(cancer, sex == 1)
mean_age <- mean(male_data$age)

# Display the result
print(mean_age)

```

```{r}
# Load the necessary library
library(survival)
library(dplyr)

# Load dataset
data(cancer, package = "survival")

# With pipes
mean_age <- cancer %>%
  filter(sex == 1) %>%
  summarise(mean_age = mean(age)) %>% 
  pull(mean_age)


# Display the result
print(mean_age)

```

## 5 **Importing data in r**
In order to work with data in R, it is important to import data from external sources such as your computer. The external data is most commonly a csv file such as Excel workbook. When we import a external data, it is converted into ``` data.frame ```.

```{r}
# Use read.csv() to import the data
# <- read.csv("C:/Users/admin/Dropbox/Cooper Project/20230509_transition_matrix/PEMBRO.OS.csv")

# Display the first few rows of the data
#head(data)

```
R also allows us to work with pre-loaded data. R comes with several datasets that are preloaded for users to explore and analyze. These datasets are part of the base R installation, and you can access them without accessing your local directory. In fact, the lung dataset we imported above is a pre-loaded dataset. 


```{r}
#Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities
library(survival)
#data function can display all available data sets in this package
data(package = "survival")
# We will use data set "lung" as an example
?data(lung)
#we can call the data set directly after loading the survival library
head(lung)
```
## 6 **Graphing using R**

```ggplot2``` is a powerful and versatile data visualization package in R that facilitates the creation of a wide range of high-quality plots. With ggplot2, you can produce a diverse array of plots, including scatter plots, bar charts, line graphs, and more, all with a focus on customization. The package encourages a declarative style, where you specify what you want the plot to look like and let ggplot2 handle the details. 


```{r, out.width="50%", warning=FALSE}
library(ggplot2)

library(survival)
data(cancer, package = "survival")
lung <- as.data.frame(cancer)

lung <- lung %>%
  mutate(sex = ifelse(sex == 1, "Male", "Female")) 

# 1. Scatter Plot of Age vs. Weight Loss
plot <-ggplot(lung, aes(x = age, y = wt.loss, color = factor(sex))) +
       geom_point() +
      labs(title = "Scatter Plot of Age vs. Weight Loss by Sex group",
       x = "Age",
       y = "Weight Loss",
       color = "Sex")
plot
# 2. Bar Chart of Gender Distribution
ggplot(lung, aes(x = factor(sex))) +
  geom_bar() +
  labs(title = "Gender Distribution",
       x = "Gender",
       y = "Count")

# 3. Boxplot of Karnofsky Performance Scores
ggplot(lung, aes(x = factor(status), y = ph.karno)) +
  geom_boxplot() +
  labs(title = "Boxplot of Karnofsky Performance Scores",
       x = "Survival Status",
       y = "Karnofsky Performance Score")


# 4. Histogram of Survival Time
ggplot(lung, aes(x = time)) +
  geom_histogram(binwidth = 50, fill = "blue", color = "black") +
  labs(title = "Histogram of Survival Time",
       x = "Survival Time",
       y = "Count")
# 5. Survival Plot

# # Load necessary libraries
library(survival)
library(survminer)

# Fit a survival model
survfit_object <- survfit(Surv(time, status) ~ 1, data = cancer)

# Create survival plot using ggsurvplot
ggsurvplot(survfit_object, data = cancer,
           xlab = "Days", ylab = "Overall survival probability",
           title = "Kaplan-Meier Survival Curve")
```


## 7 **Data Transformation**

Data transformation is the process of cleaning, organizing, and transforming raw data into a format that is easier to access and analyze. It is also known as data summarizing or data wrangling. Data transformation is important because it ensures that data is of high quality and can be used to produce accurate and valuable insights. Data analysts typically spend the majority of their time on data wrangling, as it is a necessary step before any further analysis can be done. Some useful functions under data wrangling are below. 
``` {r}

library(ggplot2)

library(survival)
data(cancer, package = "survival")
lung <- as.data.frame(cancer)


# Selecting a subset of variables
lung_selected <- lung %>%
  select(time, status, age, sex)

# Displaying the first few rows of the updated dataset
head(lung_selected)

# Generating a new variable 'sex_category' based on the 'sex' variable
lung_mutated <- lung %>%
  mutate(sex_category = ifelse(sex == 1, "Male", "Female"))

# Displaying the first few rows of the updated dataset
head(lung_mutated)


# Sorting the dataset by age in descending order
lung_sorted <- lung %>%
  arrange(desc(age))

# Displaying the first few rows of the sorted dataset
head(lung_sorted)

# Filtering the dataset to include only females
lung_females <- lung %>%
  filter(sex == 2)

# Displaying the first few rows of the filtered dataset
head(lung_females)


# Grouping by sex and calculating mean age and median survival time
summary_stats <- lung %>%
  group_by(sex) %>%
  summarize(
    MeanAge = mean(age),
    MedianSurvival = median(time)
  )

# Displaying the summary statistics
print(summary_stats)

# Grouping by sex and using slice to keep the first row in each group
lung_sliced <- lung %>%
  group_by(sex) %>%
  slice(1)  # Keep the first row in each group

# Displaying the first few rows of the sliced dataset
head(lung_sliced)
```

## 8 **Data Summary**

```{r}
# Install and load the required packages
# Load necessary libraries
library(survival)
library(dplyr)
library(summarytools)
library(stargazer)
#library(compareGroups)

# Assuming your dataset is named 'lung'
data(cancer, package = "survival")
lung <- as.data.frame(cancer)

# Use dfsummary to get a summary of the dataset
dfSummary(lung)


# Use stargazer to get a summary of the dataset
stargazer(lung, type = "text")

# Use compareGroups to get a significance of the variables in dataset with each other
#compareGroups(sex ~ ., data = lung, method = 4)

```

## 8 **Functions**

In R, a function is a reusable and self-contained block of code that performs a specific task or computation. Functions take input parameters, perform operations, and return results. They facilitate code organization, modularity, and reusability, allowing users to encapsulate logic and efficiently apply it to different data or scenarios. Functions in R are defined using the function keyword and can be called with specified arguments to execute their tasks.
```{r}
# 'cancer' is your lung cancer dataset

# Function to rescale a variable to the interval [0, 1]
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

# Example usage: Rescale the 'age' variable from the lung cancer dataset
rescaled_age <- rescale01(cancer$age)

# Display the original and rescaled values
data.frame(original_age = cancer$age, rescaled_age = rescaled_age)

```

## 9 **Hypothesis testing**


```{r}
# Load necessary libraries
library(survival)
library(dplyr)

# Assuming your dataset is named 'lung'
data(cancer, package = "survival")
lung <- as.data.frame(cancer)

# Display summary statistics
summary(lung$age)
summary(lung$time)

# Shapiro-Wilk Normality Test
# Interpretation: The Shapiro-Wilk test assesses whether the age variable follows a normal distribution. It is useful when checking the normality assumption required for certain statistical tests.
shapiro_test_result <- shapiro.test(lung$age)
print(shapiro_test_result)


# Kolmogorov-Smirnov Test for Normality
# Interpretation: The Kolmogorov-Smirnov test checks if the time variable is normally distributed. Useful for assessing normality assumption in non-parametric tests.
ks_test_result <- ks.test(lung$time, "pnorm")
print(ks_test_result)



# Q-Q Plot for Age
# Interpretation: The Q-Q plot visually assesses whether the distribution of age follows a normal distribution. If the points closely follow the diagonal line, it suggests that the age variable is approximately normally distributed.
qqnorm(lung$age)
qqline(lung$age)


# Hypothesis Testing - T-Test
# Interpretation: The t-test compares the mean age between different sexes. Useful for assessing differences in means between two groups.
# T-Test
# H0: There is no difference in mean age between sexes.
# H1: There is a significant difference in mean age between sexes
t_test_result <- t.test(age ~ sex, data = lung)
print(t_test_result)


# Hypothesis Testing - Wilcoxon Rank Sum Test
# Interpretation: The Wilcoxon Rank Sum Test evaluates if the distribution of ages differs between sexes, suitable when assumptions of normality are not met.
# Wilcoxon Rank Sum Test
# H0: There is no difference in the distribution of age between sexes.
# H1: There is a significant difference in the distribution of age between sexes.
wilcox_test_result <- wilcox.test(age ~ sex, data = lung)
print(wilcox_test_result)


# Chi-squared Test for Independence - Gender vs. Survival Status
# Interpretation: The Chi-squared test assesses the independence between gender and survival status. It is useful for analyzing associations between two categorical variables, specifically whether gender is associated with survival status.
# H0: There is no association between gender and survival status.
# H1: There is a significant association between gender and survival status.
contingency_table_gender_status <- table(lung$sex, lung$status)
contingency_table_gender_status
chi_squared_result_gender_status <- chisq.test(contingency_table_gender_status)
print(chi_squared_result_gender_status)


#The Fisher's Exact Test is a statistical test used to determine if there are nonrandom associations between two categorical variables. It is particularly useful when the sample sizes are small. This test is an exact test, meaning it calculates the exact probability of observing the given distribution of data, given the marginal totals. The test is often applied in situations where the Chi-squared test may not be appropriate due to small sample sizes or when the assumptions of the Chi-squared test are not met.

# Example using Fisher's Exact Test
# Assuming you have a 2x2 contingency table named 'my_table'
my_table <- matrix(c(10, 5, 8, 15), nrow = 2)
rownames(my_table) <- c("Group 1", "Group 2")
colnames(my_table) <- c("Category A", "Category B")

# Perform Fisher's Exact Test
fisher_test_result <- fisher.test(my_table)
print(fisher_test_result)

```


## 10 **Regression**

```{r}
# Remove rows with NA values
lung_no_na <- na.omit(lung)


# Create a correlation matrix for numeric variables
correlation_matrix <- cor(lung_no_na)

# Display the correlation matrix
print(correlation_matrix)


# Making sex variable binary
lung_no_na$sex <- ifelse(lung_no_na$sex == 1, 1, 0)


# Fit a Linear Regression (OLS) - Predicting Weight loss based on age and sex
ols_model <- lm(wt.loss ~ age + sex, data = lung_no_na)

# Display the summary of the OLS
summary(ols_model)


# Fit a Generalized Linear Model (GLM) - Predicting weight loss based on age and sex with gaussian error distribution
glm_model_gau <- glm(wt.loss ~ age + sex, data = lung_no_na, family = gaussian)

# Display the summary of the GLM
summary(glm_model_gau)


# Making censoring status variable binary
lung_no_na$status <- ifelse(lung_no_na$status == 1, 1, 0)

# Fit a Generalized Linear Model (GLM) - Predicting weight loss based on age and sex with binomial error distribution using logit
glm_model_bin <- glm(status ~ age + sex, data = lung_no_na, family = binomial(link = "logit"))

# Display the summary of the GLM
summary(glm_model_bin)

# Calculate 95% confidence intervals for coefficients
conf_intervals <- confint(glm_model_bin, level = 0.95)

# Interpretation of coefficients and odds ratios for logit
coef_summary <- summary(glm_model_bin)

# Display coefficients, odds ratios, and 95% confidence intervals
result <- cbind(
  Estimate = coef_summary$coefficients[, 1],
  Odds_Ratio = exp(coef_summary$coefficients[, 1]),
  Lower_CI = conf_intervals[, 1],
  Upper_CI = conf_intervals[, 2]
)

# Print the results
print(result)



# Fit a Quadratic Polynomial Regression  - Predicting Weight loss based on age and sex
poly_model <- lm(wt.loss ~ poly(age, degree = 2) + sex, data=lung_no_na)

# Display the summary of the poly model
summary(poly_model)

# Create a data frame for plotting
plot_data <- data.frame(age = lung_no_na$age, wt.loss = lung_no_na$wt.loss, fitted = predict(poly_model))

# Create the scatter plot with a fitted curve using ggplot2
ggplot(plot_data, aes(x = age, y = wt.loss)) +
  geom_point(color = "blue", size = 3) +
  geom_line(aes(y = fitted), color = "red", size = 1.5) +
  labs(x = "Age", y = "Weight Loss") +
  ggtitle("Quadratic Polynomial Regression") +
  theme_minimal()

#load spline package
library(splines)

# Fit a linear regression model with cubic spline for 'age'
spline_model <- lm(wt.loss ~ ns(age, df = 5) + sex, data = lung_no_na)

# Display the summary of the spline model
summary(spline_model)

# Create a data frame for plotting
# plot_data <- data.frame(age = seq(min(lung_no_na$age), max(lung_no_na$age), length.out = 100))
# 
# # Create a data frame for plotting
# plot_data <- data.frame(age = seq(min(lung_no_na$age), max(lung_no_na$age), length.out = 100))
# plot_data$fitted_spline <- predict(spline_model, newdata = data.frame(age = plot_data$age))
# 
# # Create the spline regression plot using ggplot2
# ggplot(lung_no_na, aes(x = age, y = wt.loss)) +
#   geom_point(color = "blue", size = 3) +
#   geom_line(data = plot_data, aes(x = age, y = fitted_spline), color = "green", size = 1.5) +
#   labs(x = "Age", y = "Weight Loss") +
#   ggtitle("Cubic Spline Regression") +
#   theme_minimal()



# Assuming 'cut_point' is the cutoff point for the regression discontinuity
cut_point <- 50

# Create a variable indicating whether the observation is above or below the cutoff
lung_no_na$above_cutoff <- ifelse(lung_no_na$age >= cut_point, 1, 0)

# Fit a regression discontinuity model
rd_model <- lm(wt.loss ~ above_cutoff + sex, data = lung_no_na)

#Display the summary
summary(rd_model)

# Create a data frame for plotting
plot_data <- data.frame(
  age = lung_no_na$age,
  wt.loss = lung_no_na$wt.loss,
  above_cutoff = lung_no_na$above_cutoff,
  fitted_values = predict(rd_model)
)


# Create the first ggplot graph
plot1 <- ggplot(plot_data, aes(x = age, y = fitted_values, color = factor(above_cutoff))) +
  geom_point(aes(y = wt.loss), size = 3, alpha = 0.5) +
  geom_smooth(method = 'lm') +
  labs(x = "Age", y = "Fitted Values") +
  ggtitle("Above Cutoff") +
  theme_minimal()

# Create the second ggplot graph
plot2 <- ggplot(plot_data, aes(x = age, y = wt.loss)) +
  geom_point(aes(y = wt.loss), size = 3, alpha = 0.5) +
  geom_smooth(method = 'lm') +
  labs(x = "Age", y = "Fitted Values") +
  ggtitle("Below Cutoff") +
  theme_minimal()

# Arrange the plots side by side
library(gridExtra)
grid.arrange(plot1, plot2, ncol = 2)


# Load the 'quantreg' package
# install.packages("quantreg")
library(quantreg)

# Fit quantile regression model (e.g., median, 0.5 quantile)
quantile_model <- rq(wt.loss ~ age + sex, data = lung_no_na, tau = 0.5)

# Display the summary of the quantile regression model
summary(quantile_model)

```
